Removal only

### 2020-05-26 analysis/figure-4.Rmd
better with vsd from treatment

# Figure 4. What is the effect of removal

@@

hyptsne <- addgroupings(hyptsne)
pittsne <- addgroupings(pittsne)
gontsne <- addgroupings(gontsne)
head(gontsne)

t4 <- plottsneelipsev2(pittsne, pittsne$treatment, allcolors) +  labs(subtitle = "Pituitary ") +
  facet_wrap(~sex) + theme(axis.title.x = element_blank(), strip.text = element_blank())
t5 <- plottsneelipsev2(pittsne, pittsne$extint, colorhypothesis) +  labs(subtitle = "  ")   +
  facet_wrap(~sex) + theme(axis.title = element_blank(), strip.text = element_blank())
t6 <- plottsneelipsev2(pittsne, pittsne$earlylate, colorhypothesis) +  labs(subtitle = "  ")   +
  facet_wrap(~sex) + theme(axis.title = element_blank(), strip.text = element_blank())
t456 <- plot_grid(t4, t5, t6, nrow = 1, rel_widths = c(1.1,1,1))

tsneplots <- plot_grid(t123, t456,t789, ncol = 1, rel_widths = c(1.1,1,1))
tsneplots
```

@@

```{r vsd}
vsd_path <- "../results/DEseq2/hypothesis/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

## before pivoting, check names of df with
## head(names(allvsd)) and tail(names(allvsd))

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = blk.s061.pu.y_female_gonad_inc.d9:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts") 
allvsd <- as_tibble(allvsd)

head(allvsd)
```

@@

  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1),
                  sextissue = sapply(strsplit(sextissue, '../results/DEseq2/hypothesis/'), "[", 2),
                  sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4),
                treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1),
                treatment = recode(treatment, "m.hatch" = "m.n2", "extend.hatch" = "m.n2",
                                                   "inc.prolong" = "prolong", "m.inc.d8" = "early"),
                treatment =  factor(treatment, levels = alllevels),
                earlylate = fct_collapse(treatment, 
                                  early = c("early", "m.inc.d3", "m.inc.d9", "inc.d3", 
                                         "inc.d9",  "bldg"),
                                  late = c("inc.d17",   "m.inc.d17", "prolong" ,  "hatch" , 
                                         "m.n2", "extend", "n5"),
                                  reference = "control"),
                extint = fct_collapse(treatment, 
                                  nest = c("m.inc.d3", "m.inc.d9", "m.n2", "m.inc.d17", "bldg"),
                                  eggs = c("inc.d3", "inc.d9", "inc.d17", "prolong"),
                                  chicks = c("early", "hatch", "extend", "n5"),
                                  reference = "control"))  %>%
    dplyr::select(gene, counts, sex, tissue, treatment, earlylate, extint, samples)  
  return(candidates)
}

candidatevsd <- getcandidatevsdmanip(candidategenes)
hypvsd <- candidatevsd %>% filter(tissue %in% "hypothalamus") %>% drop_na()
pitvsd <- candidatevsd %>% filter(tissue %in% "pituitary")%>% drop_na()
gonvsd <- candidatevsd %>% filter(tissue %in% "gonad")%>% drop_na()

@@

  p <- df %>%
    dplyr::filter(gene  == whichgene, treatment %in% removallevels)  %>%
    ggplot(aes(y =  counts, x = treatment, fill = treatment, color = sex)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size = 0.25, aes(color = sex)) +
    theme_B3() +
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          plot.subtitle = element_text(face = "italic")) +
    scale_color_manual(values = allcolors) +
    scale_fill_manual(values = allcolors)  +
    labs(title = paste(whichtissue, sep = " "), subtitle = whichgene, y = "gene expression") +
    #facet_wrap(~sex)  +
    geom_signif(comparisons = list(c( "inc.d3", "m.inc.d3"),
                                   c( "inc.d9", "m.inc.d9"),
                                   c( "inc.d17", "m.inc.d17"),
                                   c( "hatch", "m.n2")),
                map_signif_level=TRUE,
                textsize = 3, family = 'Helvetica',
                vjust = 1.5, size = 0.5)
  return(p)

@@

a <- plotcandidateremoval(hypvsd, "DRD1", "hypothalamus") + theme(axis.title.x = element_blank())
b <- plotcandidateremoval(pitvsd, "PRL", "pituitary")+ theme(axis.title.x = element_blank())
c <- plotcandidateremoval(gonvsd, "CRHBP", "gonads")

plot_grid(a,b,c, nrow = 3)

@@

```{r allDEGs}
DEG_path <- "../results/DEseq2/manip/"   # path to the data
DEG_files <- dir(DEG_path, pattern = "*DEGs") # get file names
DEG_pathfiles <- paste0(DEG_path, DEG_files)
#DEG_files

allDEG <- DEG_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name") %>% 
  mutate(DEG = sapply(strsplit(as.character(file_name),'./results/DEseq2/manip/'), "[", 2))  %>% 
  mutate(DEG = sapply(strsplit(as.character(DEG),'_diffexp.csv'), "[", 1))  %>% 
  mutate(tissue = sapply(strsplit(as.character(DEG),'\\.'), "[", 1)) %>%
  mutate(down = sapply(strsplit(as.character(DEG),'\\_'), "[", 3)) %>%
  mutate(up = sapply(strsplit(as.character(DEG),'\\_'), "[", 4)) %>%
  mutate(comparison = paste(down,up, sep = "_")) %>%
  mutate(sex = sapply(strsplit(as.character(sextissue),'\\_'), "[", 1)) %>%
  mutate(tissue = sapply(strsplit(as.character(sextissue),'\\_'), "[", 2)) %>%
dplyr::select(sex,tissue,comparison, direction, gene, lfc, padj, logpadj) 
head(allDEG)

allDEG$tissue <- factor(allDEG$tissue , levels = tissuelevel)
allDEG$comparison <- factor(allDEG$comparison , levels = comparisonlevels)
allDEG$direction <- factor(allDEG$direction, levels = alllevels)
head(allDEG)

candidateDEGS <- allDEG %>%
  filter(gene %in% candidategenes) %>%
  mutate(posneg = ifelse(lfc >= 0, "+", "-"),
         sex = recode(sex, "female" = "F", "male" = "M" ),
         tissue = recode(tissue, 
                         "hypothalamus" = "H",
                         "pituitary" = "P", "gonad" = "G")) %>%
  mutate(res = paste(sex, tissue, posneg, sep = "")) %>%
  select(gene, res, comparison)  %>%
  group_by(gene,  comparison) %>%
  summarize(res = str_c(res, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = res) %>%
  select(gene, contains("m.")) 
candidateDEGS

hypothesisDEGStable <- allDEG %>%
  filter(gene %in% hypothesisDEGs) %>%
  mutate(posneg = ifelse(lfc >= 0, "+", "-"),
         sex = recode(sex, "female" = "F", "male" = "M" ),
         tissue = recode(tissue, 
                         "hypothalamus" = "H",
                         "pituitary" = "P", "gonad" = "G")) %>%
  mutate(res = paste(sex, tissue, posneg, sep = "")) %>%
  select(gene, res, comparison)  %>%
  group_by(gene,  comparison) %>%
  summarize(res = str_c(res, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = res) %>%
  select(gene, contains("m.")) 
hypothesisDEGStable
```

@@

>>>>

# Figure 4. What is the effect of removal?

@@

pittsne <- addgroupings(pittsne)

t4 <- plottsneelipsev2(pittsne, pittsne$treatment, allcolors) +  labs(title = "Pituitary ", subtitle = "parental stage") +
  facet_wrap(~sex) 
t5 <- plottsneelipsev2(pittsne, pittsne$extint, colorhypothesis) +  labs(title = "  ", subtitle = "external enviornment")   +
  facet_wrap(~sex) + theme(axis.title.y = element_blank())
t6 <- plottsneelipsev2(pittsne, pittsne$earlylate, colorhypothesis) +  labs(title = "  ", subtitle = "internal timing")   +
  facet_wrap(~sex) + theme(axis.title.y = element_blank())
t456 <- plot_grid(t4, t5, t6, nrow = 1, rel_widths = c(1.1,1,1))
t456

@@

```{r vsd}
vsd_path <- "../results/DEseq2/treatment/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") 
## before pivoting, check names of df with
## head(names(allvsd)) and tail(names(allvsd))
allvsd <- allvsd %>% 
  pivot_longer(cols = L.G118_female_gonad_control:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts") 
allvsd <- as_tibble(allvsd)
head(allvsd)
tail(allvsd)

@@

  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1),
                  sextissue = sapply(strsplit(sextissue, '../results/DEseq2/treatment/'), "[", 2),
                  sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4),
                treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1),
                treatment = recode(treatment, "m.hatch" = "m.n2", "extend.hatch" = "m.n2",
                                                   "inc.prolong" = "prolong", "m.inc.d8" = "early"),
                treatment =  factor(treatment, levels = alllevels),
                earlylate = fct_collapse(treatment, 
                                  early = c("early", "m.inc.d3", "m.inc.d9", "inc.d3", 
                                         "inc.d9",  "bldg"),
                                  late = c("inc.d17",   "m.inc.d17", "prolong" ,  "hatch" , 
                                         "m.n2", "extend", "n5"),
                                  reference = "control"),
                extint = fct_collapse(treatment, 
                                  nest = c("m.inc.d3", "m.inc.d9", "m.n2", "m.inc.d17", "bldg"),
                                  eggs = c("inc.d3", "inc.d9", "inc.d17", "prolong"),
                                  chicks = c("early", "hatch", "extend", "n5"),
                                  reference = "control"))  %>%
    dplyr::select(gene, counts, sex, tissue, treatment, earlylate, extint, samples)  
  return(candidates)
}

candidatevsd <- getcandidatevsdmanip(candidategenes)
hypvsd <- candidatevsd %>% filter(tissue %in% "hypothalamus")  
pitvsd <- candidatevsd %>% filter(tissue %in% "pituitary") 
gonvsd <- candidatevsd %>% filter(tissue %in% "gonad")

# not all genes appear to be expressed in the gonads... these are
filter(gonvsd, treatment == "control") %>% head()
```

@@

  p <- df %>%
    dplyr::filter(gene  == whichgene, 
                  treatment %in% removallevels)  %>%
    ggplot(aes(y =  counts, x = treatment, fill = treatment, color = sex)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size = 0.25, aes(color = sex)) +
    theme_B3() +
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          plot.subtitle = element_text(face = "italic")) +
    scale_color_manual(values = allcolors) +
    scale_fill_manual(values = allcolors)  +
    labs(title = paste(whichtissue, sep = " "), subtitle = whichgene, y = "gene expression") +
    geom_signif(comparisons = list(c( "inc.d3", "m.inc.d3"),
                                   c( "inc.d9", "m.inc.d9"),
                                   c( "inc.d17", "m.inc.d17"),
                                   c( "hatch", "m.n2")),
                map_signif_level=TRUE,
                textsize = 3, family = 'Helvetica',
                vjust = 1.5, size = 0.5) 
  return(p)

@@

a <- plotcandidateremoval(hypvsd, "HTR2C", "hypothalamus") + theme(axis.title.x = element_blank())
b <- plotcandidateremoval(pitvsd, "PRL", "pituitary")+ theme(axis.title.x = element_blank(), strip.text = element_blank())
c <- plotcandidateremoval(gonvsd, "ESR2", "gonads") + theme(strip.text = element_blank())
boxplots1 <- plot_grid(a,b,c, nrow = 3)
boxplots1

plotcandidatetiming <- function(df, whichgene, whichtissue){

  p <- df %>%
    dplyr::filter(gene  == whichgene, 
                  treatment %in% timinglevels,
                  !treatment %in%  c("control", "bldg", "inc.d3"))  %>%
    ggplot(aes(y =  counts, x = treatment, fill = treatment, color = sex)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size = 0.25, aes(color = sex)) +
    theme_B3() +
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          plot.subtitle = element_text(face = "italic")) +
    scale_color_manual(values = allcolors) +
    scale_fill_manual(values = allcolors)  +
    labs(title = paste(whichtissue, sep = " "), subtitle = whichgene, y = "gene expression") +
    geom_signif(comparisons = list(c("inc.d9", "early"), 
                                  c( "inc.d17", "prolong"), c( "prolong", "hatch"),
                                  c( "hatch", "extend"), c( "n5", "extend")),
                map_signif_level=TRUE,
                textsize = 3, family = 'Helvetica',
                vjust = 1.5, size = 0.5) 
  return(p)

e <- plotcandidatetiming(hypvsd, "HTR2C", "hypothalamus") + theme(axis.title.x = element_blank())
f <- plotcandidatetiming(pitvsd, "PRL", "pituitary")+ theme(axis.title.x = element_blank(), strip.text = element_blank())
g <- plotcandidatetiming(gonvsd, "ESR2", "gonads") + theme(strip.text = element_blank())
boxplots2 <- plot_grid(e,f,g, nrow = 3)
boxplots2
```

```{r allDEGs}
DEG_path <- "../results/DEseq2/treatment/"   # path to the data
DEG_files <- dir(DEG_path, pattern = "*DEGs") # get file names
DEG_pathfiles <- paste0(DEG_path, DEG_files)
#DEG_files

allDEG <- DEG_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name") %>% 
  mutate(DEG = sapply(strsplit(as.character(file_name),'./results/DEseq2/treatment/'), "[", 2))  %>% 
  mutate(DEG = sapply(strsplit(as.character(DEG),'_diffexp.csv'), "[", 1))  %>% 
  mutate(tissue = sapply(strsplit(as.character(DEG),'\\.'), "[", 1)) %>%
  mutate(down = sapply(strsplit(as.character(DEG),'\\_'), "[", 3)) %>%
  mutate(up = sapply(strsplit(as.character(DEG),'\\_'), "[", 4)) %>%
  mutate(comparison = paste(down,up, sep = "_")) %>%
  mutate(sex = sapply(strsplit(as.character(sextissue),'\\_'), "[", 1)) %>%
  mutate(tissue = sapply(strsplit(as.character(sextissue),'\\_'), "[", 2)) %>%
  dplyr::select(sex,tissue,comparison, direction, gene, lfc, padj, logpadj) # %>%
 # mutate(tissue = factor(tissue, levels = tissuelevel),
  #       comparison = factor(comparison , levels = comparisonlevelsremoval),
  #       direction = factor(direction, levels = alllevels))
head(allDEG)

candidateDEGS <- allDEG %>%
  filter(gene %in% candidategenes) %>%
  mutate(posneg = ifelse(lfc >= 0, "+", "-"),
         sex = recode(sex, "female" = "F", "male" = "M" ),
         tissue = recode(tissue, 
                         "hypothalamus" = "H",
                         "pituitary" = "P", 
                         "gonad" = "G", "gonads" = "G")) %>%
  mutate(res = paste(sex, tissue, posneg, sep = "")) %>%
  select(gene, res, comparison)  %>%
  group_by(gene,  comparison) %>%
  summarize(res = str_c(res, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = res) %>%
  select(gene, inc.d3_m.inc.d3, inc.d9_m.inc.d9, inc.d17_m.inc.d17, hatch_m.n2,
          inc.d17_prolong, hatch_prolong)
candidateDEGS

@@

@@

## supplementary

```{r}

hyptsne <- addgroupings(hyptsne)
gontsne <- addgroupings(gontsne)
